---
title: "Lecture 6: Variable Transformation"
output: 
  learnr::tutorial:
    css: css/styles.css
runtime: shiny_prerendered
---

<!-- 
Before running the document, make sure `learnr` package is available. 
If missing, you can install the package with `install.packages('learnr')`.
-->

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(ggplot2)
knitr::opts_chunk$set(echo = FALSE)
set.seed(237)
obs <- sort(sample.int(30, 100, replace = TRUE))
obs_counts <- sapply(1:10, function(x) sum(obs <= (3 * x) & obs > 3 * (x - 1)))
```


## R Preparation

### Plotting histograms using `ggplot`

We will use vector `obs` to plot a histogram. The vector contains the following 100 values sorted in the ascending order.

```{r echo=TRUE}
obs
```

We can see that the values range from 1 to 30.

```{r echo=TRUE}
range(obs)
```

To plot, a histogram with bin width of 3, we can count the number of values that belong to intervals  
$$(0, 3], (3, 6], \ldots (3(n-1), 3n]\text{ for }n=1,2\ldots10.$$

```{r echo=TRUE}
obs_counts <- numeric(10)
for (i in seq_len(10)) {
  obs_counts[i] <- sum(obs <= 3 * i & obs > 3 * (i - 1))
}
obs_counts
```

Now, use `geom_bar(stat = "identity")` plot the `obs_counts` values. Also, try changing the value of `y` appropriately to update the plot to a density histogram. `obs` and `obs_counts` are available for use. 

```{r plot-hist-manual, exercise=TRUE, exercise.lines=10}
library(ggplot2)
ggplot(mapping = aes(y = obs_counts)) +
  theme_minimal() +
  geom_bar(stat = "identity")
```

```{r plot-hist-manual-hint-1}
library(ggplot2)
# histogram
ggplot(mapping = aes(x = seq_len(10) * 3, y = obs_counts)) +
  theme_minimal() +
  geom_bar(stat = "identity") +
  labs(x = "obs", y = "Counts")
```

```{r plot-hist-manual-hint-2}
library(ggplot2)
# density histogram
ggplot(mapping = aes(x = seq_len(10) * 3, y = obs_counts / (length(obs) * 3))) +
  theme_minimal() +
  geom_bar(stat = "identity") +
  labs(x = "obs", y = "Relative frequency")
```


`ggplot` provides a built-in function that plots a histogram. You can use `geom_histogram()`
and map the vector of values to count to `x` inside `aes()`. You can specify the `binwidth`
and `boundary` (where a bin begins) inside the function. Adjust the code below
to match the histograms created above.


```{r geom-hist, exercise=TRUE, exercise.lines=10}
library(ggplot2)
ggplot() +
  theme_minimal() +
  geom_histogram()
```

```{r geom-hist-hint-1}
library(ggplot2)
# histogram
ggplot(mapping = (aes(x = obs))) +
  theme_minimal() +
  geom_histogram(binwidth = 3, boundary = 0)
```

```{r geom-hist-hint-2}
library(ggplot2)
# density histogram
ggplot(mapping = (aes(x = obs, y = after_stat(density)))) +
  theme_minimal() +
  geom_histogram(binwidth = 3, boundary = 0)
```


## Generating random numbers from $U(0,1)$

+  We discussed that R provides `rxxx()` functions that generate random values from common distributions
+  We can generate from other distributions using values generated from $U(0,1)$
+  In fact, most, if not all, modern software systems including R simulate any random distributions by transforming a sequence of pseudo-random number that mimic $U(0,1)$

### Example: $U(0,1)$ to a discrete random variable

Suppose $U\sim U(0,1)$ and $X$ is a discrete random variable with probability mass function $p$.

$$p(x)=\begin{cases} 1/2 & x=0 \\ 1/3 & x = 1 \\ 1/6 & x = 3 \\ 0 & \text{otherwise}\end{cases}$$

1. Generate 100,000 samples of $U$.
2. Assign $X$ 0 if $U<1/2$, 1 if $U < 1/2 + 1/3$, and 3 otherwise.
3. Compute the relative frequency of each value of $X$ and compare with its probability mass.

```{r utodiscrete, exercise=TRUE, exercise.lines=10}
N <- 10^5
U <- runif(N) # generate samples from U(0,1)

```

```{r utodiscrete-solution}
N <- 10^5
U <- runif(N)
X <- ifelse(U < 1/2, 0, ifelse(U < 1/2 + 1/3, 1, 3))
sum(X == 0) / N
sum(X == 1) / N
sum(X == 3) / N
```


### Example: $U(0,1)$ to an exponential random variable

Suppose $U\sim U(0,1)$ and $W\sim\text{Exp}(5)$. We know that the cdf of $W$, $F_W$ is given by

$$F_W(w)=\begin{cases}1-e^{-5 w}&w>0\\0&\text{otherwise.}\end{cases}$$

1. Generate 100,000 samples of $U$.
2. Assign $W = -\log\left(1-U\right)/5$.
3. Generate 100,000 samples of $W'$ using `rexp()`.
4. Compare the density histograms of $W$ and $W'$. Use `boundary = 0`.

```{r utoexp, exercise=TRUE, exercise.lines=15}
N <- 100000
U <- runif(N) # generate samples from U(0,1)

```

```{r utoexp-solution}
N <- 100000
U <- runif(N) # generate samples from U(0,1)
W <- -log(1 - U) / 5
ggplot() +
  theme_classic() +
  geom_histogram(aes(x = W, y = after_stat(density)), boundary = 0)
```

```{r utoexp2, exercise=TRUE, exercise.lines=15}
N <- 100000
# generate samples from Exp(5) directly

```

```{r utoexp2-solution}
N <- 100000
W <- rexp(N, 5)# generate samples from Exp(5) directly
ggplot() +
  theme_classic() +
  geom_histogram(aes(x = W, y = after_stat(density)), boundary = 0)
```

### Why it works


```{r echo=FALSE, out.width="80%", fig.asp=9/16, dpi=144}
a <- 2
b <- 4
quant <- tibble(x = seq(0, 1, 0.001)) %>%
  mutate(y = pbeta(x, a, b)) %>%
  add_row(x = c(-1, 2), y = c(0, 1))
ggplot(quant, aes(x = x, y = y)) +
  theme_classic() +
  scale_y_continuous(breaks = c(0, pbeta(0.4, a, b), 1), labels = c("0", "u", "1")) +
  scale_x_continuous(breaks = c(0, 0.4, 1), labels = c("a", expression(F^{-1}*(u)), "b")) +
  annotate("segment", x = 0.4, xend = 0.4, y = 0, yend = pbeta(0.4, a, b), linetype = "dotted", linewidth = 1) +
  annotate("segment", x = -1, xend = 0.4, y =  pbeta(0.4, a, b), yend = pbeta(0.4, a, b), linetype = "dotted", linewidth = 1) +
  geom_line() +
  coord_cartesian(xlim = c(-0.2, 1.2)) +
  labs(y = "F(x)", x = "x") +
  theme(axis.text = element_text(size = 18),
        axis.title = element_text(size = 18),
        plot.title = element_text(size = 24, hjust = 0.5),
        plot.title.position = "plot")
```

+  First, note that for any number $u\in[0,1]$ and $U\sim U(0,1)$, $P(U\le u)=u$.
+  Since any cdf $F$ is between 0 and 1, $P(U\le F(x)) = F(x)$.
+  When $F$ is strictly increasing and thus invertible, we have $P(F^{-1}(U)\le x)=F(x)$.
+  This implies that $F^{-1}(U)$ follows a distribution with cdf $F$.
+  See Dekking et al. Section 6.1 and Section 6.2 for a more detailed discussion.




