---
title: "Lecture 6: Variable Transformation"
output: 
  learnr::tutorial:
    css: css/styles.css
runtime: shiny_prerendered
---

<!-- 
Before running the document, make sure `learnr` package is available. 
If missing, you can install the package with `install.packages('learnr')`.
-->

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(ggplot2)
knitr::opts_chunk$set(echo = FALSE)
```


## Generating random numbers from $U(0,1)$

+  We discussed that R provides `rxxx()` functions that generate random values from common distributions
+  We can generate from other distributions using values generated from $U(0,1)$
+  In fact, most, if not all, modern software systems including R simulate any random distributions by transforming a sequence of pseudo-random number that mimic $U(0,1)$

### Example: $U(0,1)$ to a discrete random variable

Suppose $U\sim U(0,1)$ and $X$ is a discrete random variable with probability mass function $p$.

$$p(x)=\begin{cases} 1/2 & x=0 \\ 1/3 & x = 1 \\ 1/6 & x = 3 \\ 0 & \text{otherwise}\end{cases}$$

1. Generate 100,000 samples of $U$.
2. Assign $X$ 0 if $U<1/2$, 1 if $U < 1/2 + 1/3$, and 3 otherwise.
3. Compute the relative frequency of each value of $X$ and compare with its probability mass.

```{r utodiscrete, exercise=TRUE, exercise.lines=15}
N <- 100000
U <- runif(N) # generate samples from U(0,1)

```


### Example: $U(0,1)$ to an exponential random variable

Suppose $U\sim U(0,1)$ and $W\sim\text{Exp}(5)$. We know that the cdf of $W$, $F_W$ is given by

$$F_W(w)=1-e^{-5 w}, \quad \text{for } w>0.$$

1. Generate 100,000 samples of $U$.
2. Assign $W = -\log\left(1-U\right)/5$.
3. Generate 100,000 samples of $W'$ using `rexp()`.
4. Compare the density histograms of $W$ and $W'$.

```{r utoexp, exercise=TRUE, exercise.lines=15}
N <- 100000
U <- runif(N) # generate samples from U(0,1)

```

```{r utoexp2, exercise=TRUE, exercise.lines=15}
N <- 100000
# generate samples from Exp(5) directly

```

### Why it works


```{r echo=FALSE, out.width="80%", fig.asp=9/16, dpi=144}
a <- 2
b <- 4
quant <- tibble(x = seq(0, 1, 0.001)) %>%
  mutate(y = pbeta(x, a, b)) %>%
  add_row(x = c(-1, 2), y = c(0, 1))
ggplot(quant, aes(x = x, y = y)) +
  theme_classic() +
  scale_y_continuous(breaks = c(0, pbeta(0.4, a, b), 1), labels = c("0", "u", "1")) +
  scale_x_continuous(breaks = c(0, 0.4, 1), labels = c("a", expression(F^{-1}*(u)), "b")) +
  annotate("segment", x = 0.4, xend = 0.4, y = 0, yend = pbeta(0.4, a, b), linetype = "dotted") +
  annotate("segment", x = -1, xend = 0.4, y =  pbeta(0.4, a, b), yend = pbeta(0.4, a, b), linetype = "dotted") +
  geom_line() +
  coord_cartesian(xlim = c(-0.2, 1.2)) +
  labs(y = "F(x)", x = "x") +
  theme(axis.text = element_text(size = 18),
        axis.title = element_text(size = 18),
        plot.title = element_text(size = 24, hjust = 0.5),
        plot.title.position = "plot")
```

+  First, note that for any number $y\in[0,1]$ and $U\sim U(0,1)$, $P(U\le y)=y$.
+  Since any cdf $F$ is between 0 and 1, $P(U\le F(x)) = F(x)$.
+  When $F$ is strictly increasing and thus invertible, we have $P(F^{-1}(U)\le x)=F(x)$.
+  This implies that $F^{-1}(U)$ follows a distribution with cdf $F$.
+  See Dekking et a. Section 6.1 for the general discussion.


## Simulating conditions and control flow

Suppose you and Michael play 5 rounds of Rock, Paper, and Scissors. For each round, if you win with a rock, Michael pays you \$1,  if with paper \$2, and if with scissors \$3. When you lose with the hands, you pay the assigned amounts to Michael. When you hands are the same (tie), you both put \$3 to a communal cash box. i.e., you _both_ lose \$3.

Assume both you and Michael randomly pick hand independently in each round. Let $Y$ be the amount you win after 5 rounds, $X$ be the amount that goes to the communal box, and $W$ be the number of rounds you win out of 5.

1. Simulate 100,000 plays of 5 rounds
2. For each simulation, store the amount you win, the amount that goes to the communal box, and the number of rounds you win
3. Estimate $E(Y)$, $P(Y>6)$, and $P(W\le 2)$
4. Plot a histogram of $X$

You can use the following functions to simulate a hand and compute the amount of dollors you win after each round.

```{r echo=TRUE}
play_a_hand <- function() {
  # rock is 1, paper is 2, scissors is 3
  return(sample(1:3, size = 1))
}
```

```{r rsp-setup}
play_a_hand <- function() {
  # rock is 1, paper is 2, scissors is 3
  return(sample(1:3, size = 1))
}
```

```{r rsp, exercise=TRUE, exercise.lines=45}
# example
you <- play_a_hand()
michael <- play_a_hand()
print(paste("Your hand: ", you))
print(paste("Michael's hand: ", michael))
```
